---
title: "Landscape and Patch Shape Metrics"
description: "Demo of calculating county shape and landscape metrics"
format:
  html: 
    df-print: kable
    code-fold: show
    code-summary: "Hide code"
    code-overflow: wrap
    toc-title: Page Contents
    toc: true
    toc-depth: 2
    toc-location: right
    number-sections: false
    html-math-method: katex
    smooth-scroll: true
editor: source
editor_options: 
  chunk_output_type: console
---

## Overview
This script calculates a few landscape and patch metrics as a demonstration.  In this demo, *patches* are individual Florida Counties and the landscape, or surrounding background *matrix*, is a grid.  In addition to the patches and matrix, a few points around the periphery of the couny are shown; these represent seed location for simulations of disease spread.  
     
The purpose is to show how different measures can be calculated that reflect county shape and position relative to simulations.    
   
The first half of the script walks through spatial data manipulation to extract one county polygon, rasterize it, then calculate metrics.  However, the last section uses the *calc_county_metrics()* function to iterate through all counties in Florida.         

## Libraries
```{r, warning=FALSE, message=FALSE, echo=TRUE}
library(here) #directory management
library(tidyverse) #data wrangling
library(tidyterra) # spatial
library(raster) # spatial
library(terra) # spatial
library(sf) # spatial
library(tigris) # county bounds
library(viridis) # color ramps

select <- dplyr::select # conflictwith raster
options(dplyr.summarise.inform = FALSE)
httr::set_config(httr::config(noprogress = TRUE))
options(tigris_class = "sf", tigris_use_cache = TRUE,
        tigris_progress = FALSE) # to load spatial data using sf
```

## Prepare County Data
### Get County Boundaries
Load all US Counties then subset for demo:
```{r warning=FALSE, message=FALSE}
counties_sf <- suppressMessages(counties(cb = TRUE, progress_bar=FALSE))

keep_states <- c("FL")

counties_demo <- counties(state = keep_states, cb = TRUE)

# how many counties?
length(unique(counties_demo$COUNTYFP))
```

### Project and Calc Area
```{r warning=FALSE, message=FALSE}
# US Atlas Equal Area: EPSG:9311
counties_proj <- st_transform(counties_demo, 9311)

# get crs
r_crs_sf <- crs(counties_proj)

# area in square kilometers
counties_proj <- counties_proj %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6)
```

### View Counties
```{r fig.height=10, fig.width=10}
ggplot(data = counties_proj) +
  geom_sf(aes(fill = area_km2), color = "gray40", size = 0.2) +
  scale_fill_viridis_c("Area (km)", option = "plasma") +
  theme_minimal() +
    theme(
      plot.margin = unit(c(0.1, 0.1, 0.1, 0.1), "cm"),
      legend.position = "bottom",
      legend.key.width = unit(2, "line"),
      legend.key.height = unit(1, "line"),
      legend.text = element_text(size = 10, face = "bold"),
      legend.title = element_text(size = 18, face = "bold"),
      axis.title.x = element_text(size = 18, face = "bold"),
      axis.title.y = element_text(size = 22, face = "bold"),
      axis.text.x =  element_text(size = 10, face = "bold"),
      axis.text.y = element_text(size = 10, face = "bold"),
      plot.title = element_text(size = 22, face = "bold"))
```


## Create Grid
```{r warning=FALSE, message=FALSE}
# get largest county
largest_county <- counties_proj %>%
  slice_max(area_km2, n = 1)

# which county is it?
largest_county$NAME; largest_county$STATE_NAME

# expand by 50 km
bbox <- st_bbox(largest_county)
buffer_m <- 50000 # 50 km in meters
bbox_expanded <- bbox
bbox_expanded$xmin <- bbox$xmin - buffer_m
bbox_expanded$ymin <- bbox$ymin - buffer_m
bbox_expanded$xmax <- bbox$xmax + buffer_m
bbox_expanded$ymax <- bbox$ymax + buffer_m

# create extent object
ext <- ext(bbox_expanded$xmin, bbox_expanded$xmax,
           bbox_expanded$ymin, bbox_expanded$ymax)

# grid resolution 5 km (5000 m)
res_m <- 5000

# blank raster template
r_template <- rast(ext, resolution = res_m, 
                   crs = st_crs(largest_county)$wkt)

# check
r_template
```


## Create Seed Locations
Arbitrary selection of regualrly spaced point locations around the raster.  These are intended to represent seeding locations for disease spread simulations.
```{r warning=FALSE, message=FALSE}
res_m <- res(r_template)[1] 
offset <- res_m * 2 # 2 cells, 10 km inward from edges

# bounding box
bb <- ext(r_template)
xmin <- bb[1]; xmax <- bb[2]
ymin <- bb[3]; ymax <- bb[4]

# inward offsets
xmin_in <- xmin + offset
xmax_in <- xmax - offset
ymin_in <- ymin + offset
ymax_in <- ymax - offset

# create points

# north side
x_north <- xmin_in + c(1/3, 2/3) * (xmax_in - xmin_in)
north_pts <- data.frame(x = x_north, y = ymax_in)

# south side
x_south <- xmin_in + c(1/3, 2/3) * (xmax_in - xmin_in)
south_pts <- data.frame(x = x_south, y = ymin_in)

# east side
y_east <- ymin_in + c(1/3, 2/3) * (ymax_in - ymin_in)
east_pts <- data.frame(x = xmax_in, y = y_east)

# west side
y_west <- ymin_in + c(1/3, 2/3) * (ymax_in - ymin_in)
west_pts <- data.frame(x = xmin_in, y = y_west)

# combine all
sim_origin_pnts <- rbind(north_pts, south_pts, east_pts, west_pts)

# convert to sf
sim_origin_pnts <- st_as_sf(sim_origin_pnts, coords = c("x", "y"), crs = st_crs(r_template))
```

### View Grid
Rasterize the largest county to check results. The grid was constructed based on the largest county, so all other counties should fit within it.  The yellow points represent seeding/introduction locations for disease spread simulations.
```{r fig.width=10, fig.height=10, warning=FALSE, message=FALSE}
palm_bch_fl <- rasterize(vect(largest_county), r_template, 
                        field = 1, background = 0)

grid_df <- as.data.frame(palm_bch_fl, xy = TRUE, na.rm = FALSE)
names(grid_df) <- c("x", "y", "value")

grid_df$value <- factor(grid_df$value,
                        levels = c(0, 1),
                        labels = c("Outside", "Inside"))

# as sf
largest_sf <- st_as_sf(largest_county)

ggplot() +
  geom_tile(data = grid_df, aes(x = x, y = y, fill = value)) +
  scale_fill_manual("Location", values = c("Outside" = "gray80", "Inside" = "steelblue")) +
  geom_sf(data = largest_sf, fill = NA, color = "red", linewidth = 1) +
  geom_sf(data = sim_origin_pnts, color = "black", fill = "yellow", shape = 21, size = 3, stroke = 1) +
  coord_sf(crs = st_crs(largest_sf)) +
  theme_minimal() +
    theme(
      plot.margin = unit(c(0.1, 0.1, 0.1, 0.1), "cm"),
      legend.position = "right",
      legend.key.width = unit(0.5, "line"),
      legend.key.height = unit(0.5, "line"),
      legend.text = element_text(size = 10, face = "bold"),
      legend.title = element_text(size = 18, face = "bold"),
      axis.title.x = element_text(size = 18, face = "bold"),
      axis.title.y = element_text(size = 22, face = "bold"),
      axis.text.x =  element_text(size = 10, face = "bold"),
      axis.text.y = element_text(size = 10, face = "bold"),
      plot.title = element_text(size = 22, face = "bold")
      ) + labs(title = "Palm Beach, FL \n (5km grid)")
```


## Metrics


### Patch Metrics
The *patch* refers to the geometry (area, shape, perimeter length, etc...) bounded by the county polygon.  These measures quantify patch shape and relative position in the grid.  
```{r warning=FALSE, message=FALSE}
# get county geo
geom <- sf::st_geometry(largest_county)[[1]]

# ensure it's valid
geom <- sf::st_make_valid(geom)
  
# area
area_vec_m2 <- as.numeric(sf::st_area(geom)) #m2
area_vec_km2 <- area_vec_m2 / 1e6 #to km2

# perimeter length
perim_m <- as.numeric(sf::st_length(sf::st_boundary(geom))) # meters
perim_km <- perim_m / 1000 # km

# Polsby–Popper score (range 0-1, 1 is a perfect circle)
compactness_pp <- 4 * pi * area_vec_m2 / (perim_m^2)

# shape hull area (measure of shape irregularity, 1 indicates county matches convex hull)
convex_hull_ratio <- area_vec_m2/as.numeric(sf::st_area(sf::st_convex_hull(geom)))

# shape elongation (1 is a square, values greater scale relative to increased elongation)
bb <- sf::st_bbox(geom)
bbox_w <- bb$xmax - bb$xmin
bbox_h <- bb$ymax - bb$ymin
bbox_elongation <- pmax(bbox_w, bbox_h)/pmin(bbox_w, bbox_h)

# centroid position relative to grid
# measures centrality of county within grid

#get county center
centroid <- sf::st_centroid(geom)
centroid_xy <- sf::st_coordinates(centroid)[1, ]

# get grid extent
ex <- terra::ext(palm_bch_fl) 
rel_centroid_x <- (centroid_xy[1] - ex[1]) / (ex[2] - ex[1])
rel_centroid_y <- (centroid_xy[2] - ex[3]) / (ex[4] - ex[3])
```


### Grid Metrics
Measures below based on the grid created with the county polygon (the *patch*).  The term *matrix* refers to background points, i.e., outside the county and around the *patch*. These measures look at aspects of grid and cell geometry.  
```{r warning=FALSE, message=FALSE}
# convert to raster format
r_rast <- raster::raster(palm_bch_fl)
vals <- raster::getValues(r_rast)

# 1 indicates cell is in county, everything else is NA
vals_pres <- ifelse(!is.na(vals) & vals == 1, 1L, NA_integer_)
raster::values(r_rast) <- vals_pres
  
res_m <- raster::res(r_rast)[1] # resolution in meters
cell_area_m2 <- res_m * res_m # cell area
raster_cell_count <- sum(!is.na(raster::getValues(r_rast))) # pos cell count
area_raster_km2 <- (raster_cell_count * cell_area_m2) / 1e6 # area in km

# ratio of county cels to background cells
matrix_cell_count <- sum(is.na(raster::getValues(r_rast))) 
patch_matrix_ratio <- raster_cell_count/matrix_cell_count
```   

### Simulation Origins
A few more measures.  These look at the spatial relationship between the *patch* and the locations used to initiate or seed the spread simulations.  
```{r warning=FALSE, message=FALSE}
# distances from centroid 
centroid_sfc <- sf::st_sfc(centroid, crs = r_crs_sf)
centroid_sf <- sf::st_sf(geometry = centroid_sfc)

# distances between county center and simulation seed locations
dist_to_centroid_m <- sf::st_distance(sim_origin_pnts, centroid_sf) 
dist_to_centroid_km_per_origin <- as.numeric(dist_to_centroid_m) / 1000 # to km

# minimum/closest simulation origin from above
min_dist_to_centroid_km <- min(dist_to_centroid_km_per_origin)

# distance to county boundary from simulation seed loactions
boundary_geom <- sf::st_boundary(geom)
boundary_sfc <- sf::st_sfc(boundary_geom, crs = r_crs_sf)
boundary_sf <- sf::st_sf(geometry = boundary_sfc)
dist_to_boundary_m <- sf::st_distance(sim_origin_pnts, boundary_sf) # distances
dist_to_boundary_km_per_origin <- as.numeric(dist_to_boundary_m) / 1000 # to km
min_dist_to_boundary_km <- min(dist_to_boundary_km_per_origin) # min of above
```

### Combine Metrics
Organizing data.
```{r warning=FALSE, message=FALSE}
demo_county_metrics <- tibble(
    area_vec_km2 = area_vec_km2,
    area_vec_m2 = area_vec_m2,
    perimeter_km = perim_km,
    compactness_pp = compactness_pp,
    convex_hull_ratio = convex_hull_ratio,
    bbox_elongation = bbox_elongation,
    centroid_x = centroid_xy[1],
    centroid_y = centroid_xy[2],
    centroid_rel_x = rel_centroid_x,
    centroid_rel_y = rel_centroid_y,
    raster_cell_count = raster_cell_count,
    area_raster_km2 = area_raster_km2,
    patch_matrix_ratio = patch_matrix_ratio,
    min_dist_to_centroid_km = min_dist_to_centroid_km,
    min_dist_to_boundary_km = min_dist_to_boundary_km
  )
  
# Results
data.frame(demo_county_metrics)  
```

## County Iterate
Run the same process on all counties using the *calc_county_metrics()* function.  Note: The *calc_county_metrics()* function creates points representing the simulation seed locations; this will likely need to be modified if the sim locations are pre-existing or determined through other means.

### Load Function
```{r}
source(here("2025-09-28-landscape-metrics/R/calc_county_metrics.R"))

```

### Run Function
```{r warning=FALSE, message=FALSE}
county_mets <- calc_county_metrics(counties_proj, # county polygons
                                   buffer_m = 50000, # buffer in m for fake sim points
                                   res_m = 5000) # cell resolution
```  

### Join Data
Matching metrics back to original county polygons.  
```{r}
counties_with_mets <- counties_proj %>%
  left_join(county_mets, by = c("NAME" = "county_name", "STATE_NAME" = "state_name"))

metrics_long <- counties_with_mets %>%
  st_drop_geometry() %>%
  select(NAME, STATE_NAME, # for match
         area_vec_km2, compactness_pp, # varibales to plot
         convex_hull_ratio, bbox_elongation,
         patch_matrix_ratio, min_dist_to_boundary_km
         ) %>%
  pivot_longer(
    cols = c(area_vec_km2, compactness_pp, convex_hull_ratio, bbox_elongation,
             patch_matrix_ratio, min_dist_to_boundary_km),
    names_to = "metric",
    values_to = "value"
  ) %>%
  group_by(metric) %>% # scale for a common legend
  mutate(scaled_value = (value - min(value, na.rm = TRUE)) / 
                        (max(value, na.rm = TRUE) - min(value, na.rm = TRUE))) %>%
  ungroup() %>%
  left_join(counties_proj, by = c("NAME", "STATE_NAME")) %>%
  st_as_sf()

```

### Plot Metrics
Plot a few metrics for comparison.  
```{r fig.width=12, fig.height=10}
ggplot(metrics_long) +
  geom_sf(aes(fill = scaled_value), color = "gray40", size = 0.2) +
  scale_fill_viridis_c("Value", option = "plasma") +
  facet_wrap(~ metric, ncol=3) +
  theme_minimal() +
    theme(
      plot.margin = unit(c(0.1, 0.1, 0.1, 0.1), "cm"),
      legend.position = "bottom",
      legend.key.width = unit(2, "line"),
      legend.key.height = unit(1, "line"),
      legend.text = element_text(size = 10, face = "bold"),
      legend.title = element_text(size = 18, face = "bold"),
      strip.text = element_text(size = 18, face = "bold"),
      axis.title.x = element_text(size = 18, face = "bold"),
      axis.title.y = element_text(size = 22, face = "bold"),
      axis.text.x =  element_text(size = 10, face = "bold"),
      axis.text.y = element_text(size = 10, face = "bold"),
      plot.title = element_text(size = 22, face = "bold")
      ) + labs(title = "Landscape & Patch Metrics")

```

