---
title: "Flavivirus Data Overview"
description: "Access and summarize data from secure OSF site"
format: 
   gfm:
     preview-mode: raw
editor_options: 
  chunk_output_type: console
---

## Overview
This script provides a brief demonstration on how to import data directly from an Open Science Framework project website (OSF).  Generally it's best to use GitHub for code rather than data storage; OSF provides an inexpensive (free) and secure (duel authentication) location to store the data.   

### Libraries  
Load needed libraries.    
    
The heavy lifting is performed by the [**osfr package**](https://cran.r-project.org/web/packages/osfr/index.html), which is available on CRAN, but an equivalent Python version is available from [**osfclient**](https://github.com/osfclient/osfclient).  Other packages here are just used to display imported data.  
```{r message=FALSE, warning=FALSE}
library(osfr) # OSF package
library(yaml) # yaml tools
library(tidyverse) #data wrangling
library(sf) #spatial data
library(viridis) # color palette
library(here) # directory management
```


### OSF Authentication  
Because the flavivirus detection data is stored in a secure OSF directory, a *personal access token* (PAT) is needed to access, read, and write to the project remotely.  Once a PAT is generated from your OSF profile (see: https://docs.ropensci.org/osfr/articles/auth), it can be set as an environment variable and be recognized by *osfr*, the OSF R-package.   
    
This demo is being run from a GitHub repo where my personal PAT is stored in a *secrets.yml* file.    
```{r}
# Read the PAT (must) create your own secrets.yml and PAT, this is one of mine)
my_osf_pat <- read_yaml(here("secrets.yml"))

osf_auth(my_osf_pat$osf) #authenticate to OSF
```

### Read OSF Project  
The WNV project data is located on OSF here: https://osf.io/6w7sf/ and the data is stored in the ``Disease Data`` component (sub directories are called ``components`` on OSF) of the main directory here: https://osf.io/r2z5k/    
   
**Note:**  You must have been granted access to the OSF project directories above for the hyperlinks to function properly.   
       
The last part of the data component's URL includes the GUID r2z5k, which is used to read the directory with the *osf_retrieve_node()* function.        
```{r}
wnv_project <- osf_retrieve_node("r2z5k") #Read the data  
   
#View the uploaded data files
wnv_project %>%
  osf_ls_files() #tibble with file names and formats (.sas7bat, .csv, .xlsx, etc)    
```


### Local Download  
Download files to a local directory (**Warning:** If directory is within GitHub, ensure it is included on the *.gitignore* file to avoid posting/copying the data to the repo where other might access it)  
```{r}
#Copy the file names
wnv_file_names <- wnv_project %>%
                  osf_ls_files() 

#In addition to data files, that counties_poly folder includes a US Census boundary file (ESRI shapefile) 

#download all files to a 'local/data' directory
osf_download(wnv_file_names, 
             path = here("local/data"),
             conflicts = "overwrite")
```
  
This can also be done one file at a time, in a loop or function as needed.
```{r}
#Example: just download file 5
wnv_file_names$name[5] # file name
wnv_file_names$id[5] # file id code

osf_retrieve_file(wnv_file_names$id[5]) %>% #use the id code to select files
                osf_download(conflicts = "overwrite", path = here("local/data")) #overwrite previous file 
```


### Read Example Data  
Reading the text file with human neuroinavsive West Nile Disease as an example.  
```{r}
neuro_invasive <- read_csv(here("local/data/human_neuroinvasive_wnv_2000-2021.csv"))

head(neuro_invasive)

range(neuro_invasive$year)
```

Quick bar plot
```{r fig.height=5, fig.width=8}
bar_plot <- neuro_invasive %>%
  group_by(year) %>%
  summarise(Total = sum(count, na.rm=T))

ggplot(bar_plot, aes(year, Total)) +
  geom_bar(stat="identity") +
  xlab("Year") +
  ylab("Total Cases") +
  theme_bw()
```



### Get GeoBoundary File  
This is an older, county-level shapefile (ESRI), updated ones are available here: https://www.census.gov/geographies/mapping-files/time-series/geo/carto-boundary-file.html
```{r fig.height=6, fig.width=6}
counties = read_sf(here("local/data/counties_poly/counties.shp"))

# Join 2-digit state code to 3-digit county code
counties$STCO_FIPS <- paste0(counties$STATE, counties$COUNTY) 

counties %>% 
  ggplot() +
  geom_sf() +
  theme_bw()
```

Filter to only 2020, add data to boundary file by FIPS Code, and plot 
```{r fig.height=6, fig.width=8}
cases_2020 <- neuro_invasive %>%
  filter(year == 2020) %>%  #filter data
  group_by(fips, county, state) %>%
  summarise(Count = sum(count))

counties$cases_2020 <- with(cases_2020, #match by FIPS
                        Count[match(
                           counties$STCO_FIPS,
                                    fips)])

counties$cases_2020[counties$cases_2020 == 0] = NA #setting to NA for the plot visual


ggplot() +
   geom_sf(data = counties, aes(fill = cases_2020)) + 
        scale_fill_viridis(name="Nero-Invasive WNV Cases 2020", 
                           discrete=F, option = "viridis",
                           direction = -1,
                           na.value = "white") +  
        xlab(" ") +
        ylab(" ") +
        theme(panel.grid.minor = element_blank(),
              panel.grid.major = element_blank(),
              panel.background = element_blank(),
              plot.background = element_blank(),
              panel.border = element_blank(),
              legend.position = "bottom",
              legend.title = element_text(size = 16, face = "bold"),
              axis.title.x =  element_blank(),
              axis.text.x =  element_blank(),
              axis.title.y =  element_blank(),
              axis.text.y =  element_blank(),
              axis.ticks = element_blank(),
              plot.title = element_blank())

```



